Measuring Directionality in Double Beta Decay and Neutrino Interactions with Kiloton-Scale Scintillation Detectors

### 2013-07-21 Direction_v4.tex
Version v4 of paper draft. Relative to v3 there are changes in the text, changes in Figures 1 and 4 and addition to references. Work done on July 20, 2013. --Christoph

\newcommand{\nuebar}{$\overline{\nu}_{e}$}
\newcommand{\PerTonDay}{(ton$\cdot$day)$^{-1}$}
\newcommand{\GEANT}{GEANT4}
\hyphenation{KamLAND}

@@

% All university affiliations addresses go here:
\newcommand{\ucla}{\affiliation{University of California Los Angeles, Los Angeles, CA 90095, USA}}
\newcommand{\chicago}{\affiliation{University of Chicago, Chicago, IL, 60637, USA}}

@@

\begin{abstract}
Large liquid-scintillator-based detectors have proven to be
exceptionally effective for low energy neutrino measurements due to
their good energy resolution and scalability to large volumes. The
addition of directional information using Cherenkov light and fast
timing would enhance the scientific reach of these detectors,
especially for searches for neutrino-less double-beta decay. In this
paper, we develop a technique for extracting particle direction and
evaluate different detector advances that could be used to make
direction reconstruction a reality in a kiloton-scale detector.
\end{abstract}

@@

The scalability of these detectors to large volumes also makes them
highly competitive for neutrino-less double-beta ($0\nu\beta\beta$)
decay searches in which the final state consists of a pair of
electrons with energies in the few MeV range.  Currently one of the
best limits for the $0\nu\beta\beta$ half-life comes from the scintillating detector
KamLAND-Zen\cite{KZ0nu}.

The advantage of liquid scintillators for measurements in the
$\sim$1~MeV range is their scalability from 1~ton to 1~kiloton while
providing energy resolutions of $\sim$5\%. This is roughly a factor of
two better than water Cherenkov detectors, the other developed
technology that can be economically scaled to these large
masses. However, for scintillator-based detectors, while the energy
resolution is good due to the abundance of light, the light is
isotropic and does not retain the directional information of the
primary particle.  In contrast, the direction of the particle can be
reconstructed from the Cherenkov cones in water-based detectors,
although the energy resolution rapidly degrades below $\sim$5~MeV. For
double-beta decay in particular, but also for neutrino interactions,
the directional information can be a strong suppressant of
backgrounds.

In a liquid-scintillation-based detector, Cherenkov light is also
produced, although most is absorbed and re-emitted as part of the
scintillation processes.  However, some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle or, in
the case of double beta decay, to determine the existence and topology
of the pair.  The addition of directionality is thus a powerful tool
for background rejection.  In this paper, we develop a technique for
separating the Cherenkov and scintillation light using the photon arrival
times and evaluate different detector technologies that would allow
the realization of direction reconstruction in kilo-ton scale
scintillating neutrino detectors.

\section{Light Production in Liquid Scintillators}
Liquid scintillators are cocktails of aromatic hydrocarbons. When
charged particles move through a scintillator, the molecules are
excited, predominantly via the non-localized electrons in the
$\pi$-bonds of the phenyl groups~\cite{birks_book}. Vibrational
and rotational modes of the molecules are turned into heat within
picoseconds through collisions with other molecules.  Within $\sim$10
picoseconds, the $\pi$-electrons de-excite to the first excited state
from higher levels through radiation-less transitions. The first
excited state de-excites through photon emission. There are two
characteristic times for this de-excitation, depending if the singlet
state or the triplet state was excited.  The singlet state will
de-excite within nanoseconds while the triplet state de-excites on the
order of 10's or 100's of nanoseconds. These two processes are
fluorescence and phosphorescence respectively. The exact time
constants for these processes are determined by the composition of the
scintillator.

The molecules in liquid scintillators are not isolated. Radiation-less
processes transfer energy between the molecules. The probability of
energy transfer between molecules increases as the overlap between the
molecule's absorption and emission spectra increase. The absorption
and emission spectra overlap at some level for all
molecules. Consequently if there is only one type of molecule in the
scintillator cocktail, the light output would be reduced due to
inefficiencies in the energy transfer through multiple absorption and
re-emission processes. Aromatic solutes or fluors are added to the
primary solvent to shift the wavelength of the photons to longer
wavelengths where the scintillator is more transparent. This
wavelength-shifting is also used to match the quantum efficiency as a
function of wavelength for the photodetectors being used. One typical
scintillator mixture uses pseudo-cumene as the solvent with 1-5~g/L of
PPO as the fluor. This mixture has a peak emission at about 400~nm
where bialkali photomultiplier tubes (PMTs) are most sensitive and the
pseudo-cumene is relatively transparent.

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cherenkov light will be produced as well if a particle is moving faster
than the speed of light in the medium.  This light is emitted in a
cone centered on the direction of the particle trajectory, and with a
continuous spectrum weighted toward shorter wavelengths
but extending well into the red. The spectrum is described by~\cite{Cherenkov34}:
\begin{equation}
\label{eqCherenkov}
\frac{dN}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} \left [ 1 - \frac{1}{\beta^2 n(\lambda)^2} \right ]
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the incoming particle. The Cherenkov light
produced at wavelengths shorter than the absorption cutoff of the
scintillator will be absorbed and re-emitted as isotropic light, but
wavelengths longer than this cutoff will propagate across the
detector, retaining their directional information. The yield is
roughly 60 photons per MeV, assuming a 400~nm absorption
cutoff\cite{qdot}. These undisturbed Cherenkov photons will have timing
determined by the group velocity \cite{group_velocity_article} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) - dn(\lambda)/d\textnormal{log}(\lambda)}.
\end{equation}
The longer wavelength Cherenkov photons typically arrive before
the scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelengths it should be
possible to separate the directional Cherenkov light and the isotropic
scintillation light, and then to reconstruct the direction of the
initial particle.

@@

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.25]{graphs/detector_view.png}
        \caption[]{Two back-to-back electrons with 1.41~MeV each
        (equally divided energy of $^{116}$Cd~$0\nu\beta\beta$ decay) at
        the center of the sphere with initial directions along the x
        and -x-axis. Only Cherenkov photons are drawn to illustrate the
        directionality of the event. \label{detector_view}}
        \end{center}
\end{figure}

The detector geometry is a sphere of 6.5~m diameter filled with scintillator. FIG.~\ref{detector_view} shows the geometry and the Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The default
scintillator properties have been chosen to match the KamLAND
scintillator \cite{kamland2003}: 80~\% n-Dodecane, 20~\% Pseudocumene
(1,2,4-Trimethylbenzene) and 1.52~g/l PPO (2,5-Diphenyloxazole). The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length \cite{tajimaMaster} and
refractive index \cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns) and emission decay
time constants ($\tau_{d1}$ = 6.9~ns and $\tau_{d2}$ = 8.8~ns with
relative weights of 0.87 and 0.13)
\cite{tajimaThesis}, scintillator light yield (LY, 9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)
\cite{ChrisThesis}. Variations from the baseline KamLAND case are
discussed below. Re-emission of absorbed photons in the scintillator
bulk volume and scattering have not yet been included, but are not
expected to change the conclusions here.

The inner sphere surface is used as the photodetector. It is treated
as fully absorbing (no reflections), with a photodetector coverage of
100~\%. Two important photodetector properties have been varied: 1)
the transit time spread (TTS, default $\sigma$ = 0.1~ns) and 2) the
wavelength-dependent quantum efficiency (QE) for photoelectron
production. The default is the QE of a bialkali photocathode (Hamamatsu
R7081 PMT), for which digitized values come from the Double Chooz \cite{dctwo}
Monte Carlo simulation. We note that the KamLAND 17-inch PMTs use the
same photocathode type with similar quantum efficiency.

@@

Before the simulation results for different simulation settings are
discussed, we note that  we concentrate on 5~MeV single electrons since this is the energy where scintillating detectors become more practical than water Cherenkov detectors. 
We now highlight the effects which
contribute to the timing of the scintillator detector system.  First,
the simulated travel time of the initial 5~MeV electron is between
0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the
scintillation light emission follows a distribution characterized by
scintillator-specific rise and decay times. Before the solutes in
liquid scintillator can emit optical photons, the energy has to be
transferred from the solvent to the solute. The time constant of this
energy transfer accounts for a rise time in scintillation light
emission. Past neutrino experiments were not highly sensitive to the
effect of the scintillation rise time, which is the reason why there
is a lack of accurate numbers. We assume a rise time of 1.0~ns, more
detailed studies are needed in the future. The two time constants used
to describe the falling edge of the scintillator emission time
distribution (quoted above) are values specific to the KamLAND
scintillator. Third, chromatic dispersion turns out to be an important
effect in a 6.5~m diameter detector at the level of precision needed
for direction reconstruction. Due to the wavelength-dependence of the
refractive index the speed of light in the scintillator (see Equation
(\ref{eqGroup})) increases with increasing photon wavelengths for
normal dispersion. In order to study the size of this effect, we
extracted results from a simplified simulation of 5~MeV electrons at
the center of the sphere where we used instantaneous scintillation
emission with the quantum-efficiency applied, but not including
the transit-time spread. The true hit time distributions of
photoelectrons were analyzed for scintillation light and Cherenkov
light separately. Photoelectrons coming from Cherenkov light are on
average created about 0.5~ns earlier than PEs from scintillation
light. The RMS values from PE time distributions for Cherenkov and
scintillation light are both about 0.5~ns. Note that these numbers
include the effect of the finite electron travel time.

The measurement of the arrival times of single photoelectrons is
affected by the TTS of the photodetectors, a number which can be
different by orders of magnitude depending on the detector type. The
default TTS of 0.1~ns ($\sigma$) is a value which can be achieved with
the large area picosecond photodetectors
(LAPPDs)\cite{LAPPDSum,LAPPDTDR} and possibly hybrid photodetectors
(HPDs)\cite{hpdThesis}; even significantly lower TTS numbers are
realistic with the LAPPD. A similar effect to the smearing due to the TTS, is the effect of the uncertainties in the vertex
reconstruction.

In Sections \ref{detector_timing_sec} to
\ref{scintillator_emission_sec}, we study the
photoelectron timing for different detector configurations. We focus on
the idea to increase discrimination between Cherenkov and scintillation
light by using improved detector timing. The main relevant quantities
provided by the \GEANT~simulation are the photoelectron hit positions
and the detection times after the TTS resolution has been applied. In
Section~\ref{reconstruction_sec} these quantities are then used for event
reconstruction.

@@

We first discuss results for the default simulation settings described
in the previous section. FIG.~\ref{time_plots_comparison} (a) shows
the TTS-smeared photoelectron detection times for 1000 simulated
electrons with 5~MeV energy in the center of the detector with initial
momentum directions coinciding with the x-axis. The photoelectrons
induced by Cherenkov light arrive earlier, as expected due to the
instantaneous emission and the higher average photon speed compared to scintillation light. There is however
significant overlap of the two arrival time distributions.

In order to compare simulations with different parameters, a fixed time cut $t
\leq$ 34.0~ns  is applied using the truth information to isolate the Cherenkov light in this early time window. For the default
simulation case, the number of PEs per event coming from Cherenkov
light in the early time window (108) is 98~\% of the total number of
PEs from Cherenkov light (110). For scintillation, the number of PEs (171) is only 3.2~\% of the total
scintillation-induced PEs (5445). This indicates the effectiveness of
a time cut to separate Cherenkov light from scintillation light. An
important figure of merit when comparing different simulation settings
is the ratio of Cherenkov/scintillation induced PE in the early time window since a higher ratio means more directional information per
PE. The ratio is 0.63 for the default settings.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.40]{graphs/cos_psi_34_h.pdf}
        \caption[]{The angular distribution of photoelectron hits
        relative to the original electron direction $\cos(\Psi) =
        x_{hit}/\vec{r}_{hit}$. The sample consists of 1000 events
        with a 5~MeV electron produced at the detector center. Default
        simulation settings are used and both Cherenkov and
        scintillation light are included. } 
        \label{Cherenkov_cone}
        \end{center}
\end{figure}

FIG.~\ref{Cherenkov_cone} displays the angular distribution of PE
hits after the time cut. Although this time cut is a simplification of actual time
reconstruction effects, we can use it to indicate the spatial
distribution of hits in this early time window. The Cherenkov ring structure
can be clearly seen in the peak at about 46\textdegree, demonstrating
that the directional signal conveyed by the Cherenkov photons is not
erased by scattering of the initial 5~MeV electrons.

When the 17 inch KamLAND PMTs \cite{tajimaMaster,kume_1983} (TTS =
1.28~ns) are used in the simulation, the broadening of the time
distributions leads to a strongly decreased ratio of Cherenkov over
scintillation light (0.25) for $t<34$~ns (see
FIG.~\ref{time_plots_comparison} (b)). This shows that a low
photodetector TTS is critical for directionality reconstruction and
motivates the use of novel detector types.

\section{Detector Wavelength Response}
\label{detector_wavelength_response_sec} 
In addition to improving the photodetector TTS, optimization of the
wavelength-dependent QE is promising. Since Cherenkov photons which
passed 6.5~m of scintillator have higher average wavelengths than
scintillation photons, a photodetector which is more sensitive at high
wavelengths increases not only the absolute number of PEs but also the
ratio between Cherenkov- and scintillation-induced PEs.

We have run the simulation with the QE of an extended red-sensitive
GaAsP photocathode (Hamamatsu R3809U-63 QE data was used).  FIG.~\ref{time_plots_comparison} (c) shows the results for the modified
simulation with high QE in the red spectral region. The higher
absolute number of photo-electrons coming from Cherenkov light (factor
of $\approx$ 2) and the increased Cherenkov/scintillation ratio (1.6) in the early time window would significantly improve the directionality
reconstruction.

\section{Scintillator Emission Spectrum}
\label{scintillator_emission_sec}
An alternative route towards increasing the separation in time between
Cherenkov and scintillation photon hits is the tuning of the
scintillator emission spectrum. Recently, the use of quantum dots
(QDs) in liquid scintillators has been studied as a possibility to
improve future large scale neutrino experiments \cite{qdot2}. One
major motivation for quantum-dot-doped scintillator is control of the
emission spectra by tuning the size or composition of the quantum
dots. The emission spectrum of commercial alloyed core/shell
CdS$_x$Se$_{1-x}$/ZnS quantum dots was measured. This spectrum shows a
symmetric peak centered around 461~nm with FWHM = 29~nm.

In order to isolate the effect of a different emission spectrum, the
other simulation settings, including the KamLAND absorption spectrum,
are kept unchanged. Compared to the default case shown in
FIG.~\ref{time_plots_comparison} (a) the separation is worse (as
expected) because the scintillation light wavelengths are higher than
in the KamLAND emission spectrum. However, advances in the production
of commercial quantum dot samples could yield quantum dots which have
similar, single peak emission shapes at lower wavelengths. This case
has been simulated using the same spectral shape of the measured
core-shell quantum dot emission but shifted to lower wavelengths such
that the emission peak is centered at 384~nm. This peak emission value
has been measured for other types of QDs, however with a much more
pronounced tail \cite{qdot2}. The resulting PE time distribution shows
improved separation of Cherenkov and scintillation light compared to
the default simulation. After the 34.0~ns cut on the TTS-smeared PE
time we obtain a higher Cherenkov/scintillation ratio of 0.86 (107 PE
from Cherenkov light and 124 PE from scintillation). The number of
Cherenkov-induced PEs after the time cut is unchanged while the number
of PEs coming from scintillation light is decreased due to the higher
average photon travel times.

>>>>

\newcommand{\nuebar}{$\overline{\nu}_{e}$}
\newcommand{\PerTonDay}{(ton$\cdot$day)$^{-1}$}
\newcommand{\GEANT}{GEANT4}
\newcommand{\RCS}{$R_{C/S}$}
\hyphenation{KamLAND}

@@

% All university affiliations addresses go here:
\newcommand{\ucla}{\affiliation{University of California Los Angeles, Los Angeles, CA 90095, USA}}
\newcommand{\chicago}{\affiliation{University of Chicago, Chicago, IL 60637, USA}}

@@

\begin{abstract}
Large liquid-scintillator-based detectors have proven to be
exceptionally effective for low energy neutrino measurements due to
their good energy resolution and scalability to large volumes. The
addition of directional information using Cherenkov light and fast
timing would enhance the scientific reach of these detectors,
especially for searches for neutrino-less double-beta decay. In this
paper, we develop a technique for extracting particle direction using
the difference in arrival times for Cherenkov and scintillation light,
and evaluate several detector advances in timing, photodetector
spectral response, and scintillator emission spectra that could be
used to make direction reconstruction a reality in a kiloton-scale
detector.
\end{abstract}

@@

The scalability of these detectors to large volumes also makes them
highly competitive for neutrino-less double-beta ($0\nu\beta\beta$)
decay searches in which the final state consists of a pair of
electrons with energies in the few MeV range.  The observation of this
rare decay would prove that the neutrino is a Majorana particle, which
has profound consequences to our understanding of the generation of
mass and may provide a possible explanation of the matter-antimatter
asymmetry in the universe.  Currently one of the best limits for the
$0\nu\beta\beta$ half-life comes from the scintillating detector
KamLAND-Zen\cite{KZ0nu}.

The advantage of liquid scintillators for measurements in the
$\sim$1~MeV range is their scalability from 1~ton to 1~kiloton while
providing energy resolutions of $\sim$5\%. This is roughly a factor of
two better than water Cherenkov detectors, the other developed
technology that can be economically scaled to these large
masses. However, for scintillator-based detectors, while the energy
resolution is good due to the abundance of light, the light is
isotropic and does not retain the directional information of the
primary particle.  In contrast, the direction of the particle can be
reconstructed from the Cherenkov cone in water-based detectors,
although the energy resolution rapidly degrades below $\sim$5~MeV. For
double-beta decay in particular, but also for neutrino interactions,
the directional information can be a strong suppressant of
backgrounds.

In a liquid-scintillation-based detector, Cherenkov light is also
produced, although most is absorbed and re-emitted as part of the
scintillation processes.  However, some fraction retains its
directional information. If this directional Cherenkov light can be
isolated from the copious isotropic scintillation light, it may be
possible to reconstruct the direction of the primary particle or, in
the case of double beta decay, to determine the existence and topology
of the pair.  The addition of directionality is thus a powerful tool
for background rejection.  In this paper, we develop a technique for
separating the Cherenkov and scintillation light using the photon arrival
times and evaluate several detector technologies that would allow
the realization of direction reconstruction in kilo-ton scale
scintillating neutrino detectors.

\section{Liquid Scintillator Detectors}
Liquid scintillators are `cocktails' of aromatic hydrocarbons. When
charged particles move through a scintillator, the molecules are
excited, predominantly via the non-localized electrons in the
$\pi$-bonds of the phenyl groups~\cite{birks_book}. Vibrational
and rotational modes of the molecules are turned into heat within
picoseconds through collisions with other molecules.  Within $\sim$10
picoseconds, the $\pi$-electrons de-excite to the first excited state
from higher levels through radiation-less transitions. The first
excited state de-excites through photon emission. There are two
characteristic times for this de-excitation, depending if the singlet
state or the triplet state was excited.  The singlet state will
de-excite within nanoseconds while the triplet state de-excites on the
order of 10's or 100's of nanoseconds. These two processes are
fluorescence and phosphorescence respectively. The exact time
constants for these processes are determined by the composition of the
scintillator.

The absorption and emission spectra overlap at some level
in all molecules. Consequently, if there is only one type of molecule in
the scintillator cocktail the light output is reduced due to
inefficiencies in the energy transfer through multiple absorption and
re-emission processes. Aromatic solutes or fluors are added to the
primary solvent to shift the wavelength of the photons to longer
wavelengths where the scintillator is more transparent. This
wavelength-shifting is also used to match the quantum efficiency as a
function of wavelength for the photodetectors being used. One typical
scintillator mixture uses pseudo-cumene as the solvent with 1-5~g/L of
PPO as the fluor. This mixture has a peak emission at about 400~nm
where bialkali photomultiplier tubes (PMTs) are most sensitive and the
pseudo-cumene is relatively transparent.

A good liquid scintillator will produce $\sim$10,000 photons
isotropically per MeV of deposited energy. Although less abundant,
Cherenkov light will be produced as well if a particle is moving
faster than the speed of light in the medium.  This light is emitted
in a cone centered on the direction of the particle trajectory, and
with a continuous spectrum weighted toward shorter wavelengths but
extending well into the red. The spectrum is described
by~\cite{Cherenkov34}:
\begin{equation}
\label{eqCherenkov}
\frac{dN}{d\lambda dx} = \frac{2 \pi \alpha Z^2}{\lambda^2} \left [ 1 - \frac{1}{\beta^2 n(\lambda)^2} \right ]
\end{equation}
where $n(\lambda)$ is the wavelength-dependent index of refraction and
$\beta$ is the velocity of the incoming particle. In a large detector,
the Cherenkov light produced at wavelengths shorter than the
absorption cutoff of the scintillator will be absorbed and re-emitted
as isotropic light, but wavelengths longer than this cutoff will
propagate across the detector, retaining their directional
information. The yield is roughly 60 photons per MeV, assuming a
400~nm absorption cutoff\cite{qdot}. These undisturbed Cherenkov
photons will have timing determined by the group velocity
\cite{group_velocity_article} in the liquid,
\begin{equation}
\label{eqGroup}
v_{g}(\lambda) = \frac{c_{vacuum}}{n(\lambda) - dn(\lambda)/d\textnormal{log}(\lambda)}.
\end{equation}
The longer wavelength Cherenkov photons typically arrive before the
scintillation light, which is slowed by both the scintillation
processes and the shorter wavelengths involved. Thus, with sufficient
timing resolution and sensitivity to longer wavelengths it should be
possible to separate the directional Cherenkov light and the isotropic
scintillation light, and then to reconstruct the direction of the
initial particle.

In $0\nu\beta\beta$, the electrons emerge with a combined energy equal
to the Q-value of the particular isotope. The individual
electrons come out following distributions of energies and angular correlations, 
a probable case being equal division of energy between back-to-back
electrons\cite{phasespace}. This case is shown in FIG.~\ref{detector_view}
for an example $^{116}$Cd $0\nu\beta\beta$ event. Since the decay
half-life is inversely proportional to the phase-space factor, isotopes with
higher Q-values are preferred and due to backgrounds from the daughters
of the $^{238}$U and $^{232}$Th decay chains those with Q-values at or
above 2.6~MeV are most often considered for $0\nu\beta\beta$ searches.
There are hundreds of candidate $0\nu\beta\beta$ isotopes\cite{tabledbb},
but only a handful with large Q-values.  Most of the high Q-value candidates
have been considered as a dopant for a liquid scintillator:
$^{150}$Nd(Q=3.367~MeV)\cite{minfang,nd1}, $^{96}$Zr(Q=3.350~MeV)\cite{zr1},
$^{100}$Mo(Q=3.034~MeV)\cite{mo1}, $^{82}$Se(Q=3.350~MeV)\cite{qdot},
$^{116}$Cd(Q=2.802~MeV)\cite{qdot, cd1}, $^{130}$Te(Q=2.533~MeV)\cite{qdot, biller},
$^{136}$Xe(Q=2.479~MeV)\cite{KZ0nu} and $^{124}$Sn(Q=2.29~MeV)\cite{sn1}.
Xenon gas readily dissolves into liquid scintillator. For the other isotopes,
suitable organometallic compound needs to be found that produces a stable
scintillator with a long attenuation length in the wavelength region of
interest. Recently, nanocrystals called quantum dots have been explored
as an alternative method for doping scintillator\cite{qdot,qdot2}. Quantum dots
have interesting optical properties which are discussed further
in Section~\ref{scintillator_emission_sec}.

The reconstruction of the direction of the two simultaneous
emitted $\sim$1.5~MeV electrons from the $0\nu\beta\beta$ decay is
quite challenging. In this paper, we decided to start our directional
studies with 5~MeV single electrons since reconstruction of direction
in this energy range had never been achieved. This is a logical
energy to use since this is where historically scintillating detectors
become more practical than water Cherenkov detectors. This is also
an energy that is relevant for studies of $^{8}$B solar neutrinos
and neutrino-electron elastic scattering measurement.

@@

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.4]{graphs/geometry_plot_labels.pdf}
        \caption[]{Two back-to-back electrons with 1.41~MeV each
        (equally divided energy of $^{116}$Cd~$0\nu\beta\beta$ decay) at
        the center of the sphere with initial directions along the x
        and -x-axis. Only Cherenkov photons (green lines) are drawn to illustrate the
        directionality of the event. \label{detector_view}}
        \end{center}
\end{figure}

The detector geometry is a sphere of 6.5~m radius filled with
scintillator. FIG.~\ref{detector_view} shows the geometry and the
Cherenkov light from an example $^{116}$Cd $0\nu\beta\beta$ event. The
default scintillator properties have been chosen to match the KamLAND
scintillator \cite{kamland2003}: 80\% n-Dodecane, 20\% Pseudocumene
(1,2,4-Trimethylbenzene) and 1.52~g/l PPO (2,5-Diphenyloxazole). The
scintillator properties implemented in the simulation include the
atomic composition and density ($\rho$ = 0.78~g/ml), the
wavelength-dependent attenuation length \cite{tajimaMaster} and
refractive index \cite{OlegThesis}, the scintillation emission
spectrum\cite{tajimaMaster}, emission rise time ($\tau_r$ = 1.0~ns)
and emission decay time constants ($\tau_{d1}$ = 6.9~ns and
$\tau_{d2}$ = 8.8~ns with relative weights of 0.87 and 0.13)
\cite{tajimaThesis}, scintillator light yield (LY, 9030 photons/MeV)
and the Birks constant ($kB$ $\approx$ 0.1~mm/MeV)
\cite{ChrisThesis}. Variations from the baseline KamLAND case are
discussed below. Re-emission of absorbed photons in the scintillator
bulk volume and scattering have not yet been included, but are not
expected to change the conclusions here.

The inner sphere surface is used as the photodetector. It is treated
as fully absorbing (no reflections), with a photodetector coverage of
100\%. Two important photodetector properties have been varied: 1)
the transit time spread (TTS, default $\sigma$ = 0.1~ns) and 2) the
wavelength-dependent quantum efficiency (QE) for photoelectron
production. The default is the QE of a bialkali photocathode (Hamamatsu
R7081 PMT)\cite{Hamamatsu_R7081}, for which digitized values come from the Double Chooz \cite{dctwo}
Monte Carlo simulation. We note that the KamLAND 17-inch PMTs use the
same photocathode type with similar quantum efficiency.

@@

Before the simulation results for different simulation settings are
discussed, we note that we concentrate on 5~MeV single electrons since
this is the energy at which scintillating detectors become more practical
than water Cherenkov detectors.  We now highlight the effects which
contribute to the timing of the scintillator detector system.  First,
the simulated travel time of the initial 5~MeV electron is between
0.10 and 0.15~ns, while the travel distance is about 3~cm. Second, the
scintillation light emission follows a distribution characterized by
scintillator-specific rise and decay times. Before the solutes in
liquid scintillator can emit optical photons, the energy has to be
transferred from the solvent to the solute. The time constant of this
energy transfer accounts for a rise time in scintillation light
emission. Past neutrino experiments were not highly sensitive to the
effect of the scintillation rise time, which is the reason why there
is a lack of accurate numbers. We assume a rise time of 1.0~ns; more
detailed studies are needed in the future. The two time constants used
to describe the falling edge of the scintillator emission time
distribution (quoted above) are values specific to the KamLAND
scintillator. Third, chromatic dispersion turns out to be an important
effect in a 6.5m-radius detector at the level of precision needed for
direction reconstruction.

Due to the wavelength-dependence of the refractive index the speed of
light in the scintillator (see Equation (\ref{eqGroup})) increases
with increasing photon wavelengths for normal dispersion, with red
light traveling faster than blue light.  In order to study the
measurability of the time differences, we simulated 5~MeV electrons at
the center of the sphere where we used instantaneous scintillation
emission with the quantum-efficiency applied, but not including a
transit-time spread. The true hit time distributions of photoelectrons
were analyzed for scintillation light and Cherenkov light
separately. Photoelectrons coming from Cherenkov light are on average
created about 0.5~ns earlier than PEs from scintillation light. The
RMS values from PE time distributions for Cherenkov and scintillation
light are both about 0.5~ns. Note that these numbers include the
effect of the finite electron travel time.

The measurement of the arrival times of single photoelectrons is
affected by the transit-time spread (TTS) of the photodetectors, a
number which can be different by orders of magnitude depending on the
detector type. The default TTS of 0.1~ns ($\sigma$) is a value which
can be achieved with the large area picosecond photodetectors
(LAPPDs)\cite{LAPPDSum,LAPPDTDR} and possibly hybrid photodetectors
(HPDs)\cite{hpdThesis}; even significantly lower TTS numbers are
realistic with the LAPPD~\cite{RSI_paper,PSEC4_paper,anode_paper}. We
note that uncertainties in the vertex reconstruction.will produce a
similar effect to the smearing due to the TTS.

In Sections \ref{detector_timing_sec} to
\ref{scintillator_emission_sec}, we study the
photoelectron timing for different detector configurations. We focus
on the idea to increase discrimination between Cherenkov and
scintillation light by using improved detector timing. The primary
quantities provided by the \GEANT~simulation are the photoelectron hit
positions and the detection times after the TTS resolution has been
applied. In Section~\ref{reconstruction_sec} these quantities are then
used for event reconstruction.

@@

We first discuss results for the default simulation settings described
in the previous section. FIG.~\ref{time_plots_comparison} (a) shows
the TTS-smeared photoelectron (PE) detection times for 1000 simulated
electrons with 5~MeV energy in the center of the detector with initial
momentum directions coinciding with the x-axis. The photoelectrons
induced by Cherenkov light arrive earlier, as expected due to the
instantaneous emission and the higher average photon speed compared to
scintillation light. There is however significant overlap of the two
arrival time distributions.

In order to compare simulations with different parameters, a fixed
time cut of $t\leq$ 34.0~ns is applied using the truth information to
isolate the Cherenkov light in this early time window. For the default
simulation case, the average number of PEs per event coming from
Cherenkov light in the early time window (108) is 98\% of the total
average number of PEs from Cherenkov light (110). For scintillation
light, the average number of PEs (171) is only 3.2\% of the average total
scintillation-induced PEs (5445). This indicates the effectiveness of
a time cut to separate Cherenkov light from scintillation light.

The ratio of Cherenkov-induced to scintillation-induced photo-electrons
in the early time window (\RCS) is a useful figure-of-merit when
comparing different simulation settings, since a higher ratio means
more directional information per PE. For the default simulation
settings $R_{C/S}=0.63$.

\begin{figure}
        \begin{center}
        \includegraphics[scale=0.40]{graphs/cos_psi_34_h.pdf}
        \caption[]{The angular distribution of photoelectron hits
        relative to the original electron direction $\cos(\Psi) =
        x_{hit}/\vec{r}_{hit}$. The sample consists of 1000 events
        with a 5~MeV electron produced at the detector center. Default
        simulation settings are used and both Cherenkov and
        scintillation light are included.} 
        \label{Cherenkov_cone}
        \end{center}
\end{figure}

FIG.~\ref{Cherenkov_cone} displays the angular distribution of PE
hits after the time cut. Although this time cut is a simplification of actual time
reconstruction effects, we can use it to indicate the spatial
distribution of hits in the early time window. The Cherenkov ring structure
can be clearly seen in the peak near 46\textdegree, demonstrating
that the directional signal conveyed by the Cherenkov photons is not
erased by scattering of the initial 5~MeV electrons.

When the 17-inch KamLAND PMTs \cite{tajimaMaster,kume_1983} (TTS =
1.28~ns) are used in the simulation, the broadening of the time
distributions leads to a strongly decreased ratio of Cherenkov over
scintillation light ($R_{C/S}=0.25$) for $t<34$~ns (see
FIG.~\ref{time_plots_comparison} (b)). This shows that a low
photodetector TTS is critical for directionality reconstruction and
motivates the use of novel detector types.

\section{Detector Wavelength Response}
\label{detector_wavelength_response_sec} 
In addition to decreasing the photodetector TTS to enhance $R_{C/S}$,
it is possible to optimize the wavelength-dependence of the
photocathode. Since Cherenkov photons which passed through meters of
scintillator have longer average wavelengths than scintillation
photons, a photodetector which is more sensitive at long wavelengths
increases not only the absolute number of PEs but also the ratio
between Cherenkov- and scintillation-induced PEs.

We have run the simulation with the QE of an extended red-sensitive
GaAsP photocathode (Hamamatsu R3809U-63)~\cite{Hamamatsu_R3899U}.
FIG.~\ref{time_plots_comparison}(c) shows the results for the
modified simulation with high QE in the red spectral region. The
higher absolute number of photo-electrons coming from Cherenkov light
(factor of $\approx$ 2) and the increased Cherenkov/scintillation
ratio (1.6) in the early time window would significantly improve the
directionality reconstruction.
% Can we quote a value of R for this? hjf

\section{Scintillator Emission Spectrum}
\label{scintillator_emission_sec}
An alternative route towards increasing the separation in time between
Cherenkov and scintillation photon hits is the tuning of the
scintillator emission spectrum. Recently, the use of quantum dots
(QDs) in liquid scintillators has been studied as a possibility to
improve future large scale neutrino experiments~\cite{qdot2}. One
major motivation for quantum-dot-doped scintillator is control of the
emission spectra by tuning the size or composition of the quantum
dots.

The emission spectrum of commercial alloyed core/shell
CdS$_x$Se$_{1-x}$/ZnS quantum dots was measured in
Ref.~\cite{qdot2}. This spectrum shows a symmetric peak centered
around 461~nm with FWHM = 29~nm.  In order to isolate the effect of
the different emission spectrum, the other simulation settings,
including the KamLAND absorption spectrum, were kept unchanged; we
find $R_{C/S}=0.17$ for the default 34-nsec timing cut.  Compared to
the default case shown in FIG.~\ref{time_plots_comparison}(a) the
separation is worse (as expected) because the scintillation light
wavelengths are higher than in the KamLAND emission spectrum.

However, advances in the production of commercial quantum dot samples
could yield quantum dots which have similar, single peak emission
shapes at lower wavelengths. This case has been simulated using the
same spectral shape of the measured core-shell quantum dot emission
but shifted to lower wavelengths such that the emission peak is
centered at 384~nm. This peak emission value has been measured for
other types of QDs, however with a much more pronounced tail
\cite{qdot2}. The resulting PE time distribution shows improved
separation of Cherenkov and scintillation light compared to the
default simulation. After the 34.0~ns cut on the TTS-smeared PE time
we obtain a Cherenkov/scintillation ratio of $R_{C/S}=0.86$ (107 PE
from Cherenkov light and 124 PE from scintillation). The number of
Cherenkov-induced PEs after the time cut is unchanged while the number
of PEs coming from scintillation light is decreased due to the higher
average photon travel times.
